---
title: "DA101 - Final project assignment"
author: "Wilson Le"
date: "10/21/2020"
output: 
  html_document:
    code_folding: hide
---

THE EFFECTS OF CLIMATE ON THE SPREAD OF COVID-19
===

## Introduction:

Corona virus disease 2019 (COVID-19), initially identified in December 2019 in Wuhan, China, has now become a global pandemic. It is a very infectious virus that spreads between humans, mainly via physical contact with bodily fluids[1]. Scientific research has shown that climate temperature plays a role in the spreading of the virus[2]. This research aim to redo the experiment using different data and techniques to see if the claims made in the researches are true.

Research will try to answer the question: Does climate temperature affects the spread of Covid-19? And how? The dependent variable, which will be the variable used to determine the "spread of Covid-19", will be the daily confirmed cases of a country. By looking at a single country, the data will be more uniform as different country has different population size and different seriousness when it comes to response to the pandemic, which consequently affect the number of confirmed cases. This variable will be extracted from the Covid-19 data by Out world in Data[4]. The independent variable that might effect the spread of Covid-19 that the research is going to look at is minimum and maximum temperature of different regions in a country (i.e different states in the US). Temperature data set of the US states since January to October[5] will be used to extract the necessary data to perform analysis. The topic question will be an exploratory type of question, as it will be looking for unknown patterns within the data.

## Ethical Considerations:

Since the researching the research is using data from the internet for analysis, the data itself are public for every one to look at and process. However, the research is conducted by an individual (me), however carefully conducted, there will be flaws and mistakes in logic and assumptions. There are potential for harm in this analysis. As Covid19 is a very unpredictable infectious disease, these analysis should only be part of the consideration of your decision to act upon the disease.

## Data processing

Import libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(GGally)
library(stringi)
```

<br/><br/>

Import data:
```{r}
Covid19Data <- read.csv('Covid-19-All-State-History.csv')
maxTempData <- read.csv('./csv/tMax_US.csv')
minTempData <- read.csv('./csv/tMin_US.csv')
windData <- read.csv('./csv/wind_US.csv')
stateData <- read.csv('./csv/states.csv')
```

<br/><br/>

Process Covid19 data:


```{r}
Covid19Data <- Covid19Data %>%
  
  # Select the wanted variables:
  select(date, state, positiveCasesViral, totalTestsPeopleViral) %>%
  
  # Remove incomplete observations:
  na.omit()
  
# Aggregate monthly sum positive case of each state:
SumPositiveCase <- aggregate(Covid19Data$positiveCasesViral, by=list(state=Covid19Data$state, date = Covid19Data$date), FUN=sum) %>% rename(positiveCaseViral = x)

# Aggregate monthly sum tests of each state:
SumTests <- aggregate(Covid19Data$totalTestsPeopleViral, by=list(state=Covid19Data$state, date = Covid19Data$date), FUN=sum) %>% rename(totalTestsPeopleViral = x)

# Merge the two together:
Covid19SumData <- merge(SumPositiveCase, SumTests, by=c("state", "date"))


kable(head(Covid19SumData, n=20))
```

<br/><br/>

Process max temperature data:
```{r}
# Calculate the mean maximum daily temperature.
maxTempbyProv <- aggregate(maxTempData[12:269] , by=list(province=maxTempData$Province_State), FUN=mean)


# Select just the state and the mean maximum daily temperature of states.
maxTempbyProv <- maxTempbyProv %>% select(province, starts_with("X"))

# Define function to transpose a state's daily temperature data.
transposeMaxState <- function(stateName){
  df <- maxTempbyProv %>% filter(province == as.character(stateName)) %>%
    select(starts_with("X")) %>%
    t() %>%
    as.data.frame()
  df <- df %>% 
    mutate(unformatedDate = substr(rownames(df), start=2, stop=8)) %>% 
    mutate(province = stateName) %>% rename(Max = V1) %>% 
    mutate(date = as.Date.character(unformatedDate, format="%m.%d.%y")) %>%
    select(!unformatedDate)
  return(df)
}

# Define function to bind all the transposed data.
bindStates <- function(){
  bindDf <- data.frame()
  for (state in maxTempbyProv$province){
    tempDf <- transposeMaxState(state)
    bindDf <- rbind(bindDf, tempDf)
  }
  return(bindDf)
}

# Transposed maximum temperature data
maxTempbyProvTransposed <- bindStates()

#Display the first 10 value of the new data set.
kable(head(maxTempbyProvTransposed, n=20))
```

<br/><br/>

Process min temperature data:
```{r}
#Calculate the mean minimum daily temperature of a state.
minTempbyProv <- aggregate(minTempData[12:269] , by=list(province=minTempData$Province_State), FUN=mean)

# Select just the state and the mean maximum daily temperature of states.
minTempbyProv <- minTempbyProv %>% select(province, starts_with("X"))

# Define function to transpose a state's temperature per month data.
transposeMinState <- function(stateName){
  df <- minTempbyProv %>% filter(province == as.character(stateName)) %>%
    select(starts_with("X")) %>%
    t() %>%
    as.data.frame()
  df <- df %>% 
    mutate(unformatedDate = substr(rownames(df), start=2, stop=8)) %>% 
    mutate(province = stateName) %>% rename(Min = V1) %>% 
    mutate(date = as.Date.character(unformatedDate, format="%m.%d.%y")) %>%
    select(!unformatedDate)
  return(df)
}

# Define function to bind all the transposed data.
bindStates <- function(){
  bindDf <- data.frame()
  for (state in minTempbyProv$province){
    tempDf <- transposeMinState(state)
    bindDf <- rbind(bindDf, tempDf)
  }
  return(bindDf)
}

# Transposed maximum temperature data
minTempbyProvTransposed <- bindStates()

#Display the first 10 value of the new data set.
kable(head(minTempbyProvTransposed, n=20))
```

<br/><br/>

Process wind data:
```{r}
#Filter out invalid values:
windDataFiltered <- windData %>% filter_at(vars(starts_with("X")), all_vars(. > 0))

#Calculate the mean minimum daily wind speed of a state.
windByProv <- aggregate(windDataFiltered[12:269] , by=list(province=windDataFiltered$Province_State), FUN=mean)

# Select just the state and the mean maximum daily wind speed of states.
windByProv <- windByProv %>% select(province, starts_with("X"))

# Define function to transpose a state's temperature per month data.
transposeWinState <- function(stateName){
  df <- windByProv %>% filter(province == as.character(stateName)) %>%
    select(starts_with("X")) %>%
    t() %>%
    as.data.frame()
  df <- df %>% 
    mutate(unformatedDate = substr(rownames(df), start=2, stop=8)) %>% 
    mutate(province = stateName) %>% rename(Wind = V1) %>% 
    mutate(date = as.Date.character(unformatedDate, format="%m.%d.%y")) %>%
    select(!unformatedDate)
  return(df)
}

# Define function to bind all the transposed data.
bindStates <- function(){
  bindDf <- data.frame()
  for (state in windByProv$province){
    tempDf <- transposeWinState(state)
    bindDf <- rbind(bindDf, tempDf)
  }
  return(bindDf)
}

# Transposed maximum temperature data
windByProvTransposed <- bindStates()

#Display the first 10 value of the new data set.
kable(head(windByProvTransposed, n=20))
```

<br/><br/>

Merge temperature and wind data:

```{r warning=FALSE, message=FALSE}
# Merge the max and min temperature into one data set.
TempByProv <- merge(maxTempbyProvTransposed, minTempbyProvTransposed, by = c("province", "date"))
ClimateByProv <- merge(TempByProv, windByProvTransposed, by = c("province", "date"))
# Display the first 10 values of the merged temperature data set.
options(digits = 3)
kable(head(ClimateByProv, n = 10))
```

<br/><br/>

Merge temperature data with Covid19 data:
```{r warning=FALSE, message=FALSE}
ClimateByState <- merge(ClimateByProv, stateData, by=c("province"))
Covid19Climate <- merge(ClimateByState, Covid19SumData, by=c("state", "date"))

# Display the first 10 values of the merged temperature data set.
options(digits = 3)
kable(head(Covid19Climate, n = 10))
```


## Data Explanation and Exploration:

As stated above, the research will be looking at the spread of Covid-19, which is determined by the number of confirmed positive test cases in the states. The variables that might affect the spread are the temperature of the states.

The dependent variable will be taken from the COVID tracking project’s data set. The project is a non-profit organization tasked to collect and publish data required to quantify the spread and seriousness of COVID-19 in the United States. The data is now being collected by volunteers and processed and analyzed by developers, scientists, designers, editors, reporters, and many enthusiast contributors[5]. Despite being collected by volunteers, there are strict quality scoring methods in place for all these collected data. This project has been collecting data since January of 2020. With the vast amount of data and its precision, this data set samples the real situation of any collected regions. There are many variables in this data set, but this research only pays attention to a few of them. The “date” and “state” are 2 necessary variables to tell when and where the data is collected, along with the number that we are looking for, the “positiveCasesViral” and “totalTestsPeopleViral”. According to the metadata provided by the data owners, the variable “positiveCasesViral” corresponds to “the total number of unique people with a complete PCR test that returns positive”, or in other words, the total of confirmed cases, and the variable “totalTestsPeopleViral” corresponds to the total number of “unique people tested at least once via PCR testings”[6]. These 4 variables from this data set will provide data of confirmed cases in each state over a time period, which consequently corresponds to the spread of the pandemic in state areas of the whole country.

Moving on to the two independent variables, the data set collected by Eeemonts on Kaggle will provide precise data. This data set is actually compiled by calling API requests from Dark Sky, a company that’s recently acquired by Apple but the API data collection is still available. This company measures weather information and updates it by the minute[7]. Eeemonts has been pulling and compiling data since January 2020, resulting in a massive cluster of data sets that includes a multitude of variables[5]. The sample that this cluster of data sets is a detailed data weather information of every city of state every day since January. However, we are only paying attention to two data sets, the “tMax_US.csv” and “tMin_US.csv”. These data sets include two important variables that the research needs: the maximum temperature and the minimum temperature of a city in a state in a day, which, will provide the range of temperature of a state.

The challenge is that these two data set are formatted differently, so data wrangling is a challenge step, in order to produce a neat and usable data set for analysis. The Covid19 Data has more rows than columns, as the "Date" and "State" variable has its own column, meaning that for the number of rows will be the product of the number of states and the number of dates, resulting in more rows than columns. As for the temperature data set, there is more columns than rows. This is because instead of having a column specifically for dates, the data set is formatted so that every day has its own columns, and the intersection between the date and the state will be the temperature value. This means that the number of rows will be equal to the number of states, but the number of columns will be as many as the number of days, resulting in a data that has more columns than rows. To deal with this is a challange, as I have to transpose the temperature data and wrote my own function to turn it into the same format as the Covid-19 data. Only then can I merge the two data set together and start with my data exploration and analysis.

Bar plot showing the change in __number of positive test cases.__ in the entire US in 2020.
```{r}
#Aggregate monthly sum test of entire US:
Covid19SumPos <- aggregate(Covid19Climate$positiveCaseViral, by=list(date = Covid19Climate$date), FUN=sum) %>% rename(positiveCaseViral = x)

#Plot
ggplot(Covid19SumPos, aes(date, positiveCaseViral)) + 
  geom_line() +
  labs(
    title="Covid19 positive test cases",
    caption='The change in Covid19 positive test cases in the entire US in 2020',
    x="Time",
    y="Positive test cases") + 
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
```

Frequency polygon showing distribution of maximum and minimum temperature across all the states:
```{r}
ggplot(Covid19Climate) + 
  geom_freqpoly(bins = 40, aes(Max), color="red", size=3) +
  geom_freqpoly(bins = 40, aes(Min), color="steelblue", size=3) +
  labs(title="Maximum (red) and Minimum (blue) temperature across the states",
    caption="Frequency polygon representing distribution of maximum (red) and minimum (blue) temperature.",
    x="degrees Celcius",
    y="Count") + 
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
```

Line plot showing changes in maximum and minimum temperature across the months:
```{r}
#Calculate daily mean maximum temperature across all state.
Covid19MaxTemp <- aggregate(Covid19Climate$Max, by=list(date=Covid19Climate$date), FUN=mean) %>%
  rename(Max = x)

#Calculate daily mean minimum temperature across all state.
Covid19MinTemp <- aggregate(Covid19Climate$Min, by=list(date=Covid19Climate$date), FUN=mean) %>%
  rename(Min = x)

#Merge the two to create a data frame of daily mean minimum, maximum temperaturea across all states.
Covid19DailyTemp <- merge(Covid19MaxTemp, Covid19MinTemp, by="date")

ggplot(Covid19DailyTemp) + 
  geom_line(aes(date, Max), color="red") +
  geom_line(aes(date, Min), color="blue") +
  labs(
    title="Change of max and min temperature", 
    caption="Change in daily maximum (red) and minimum (blue) temperature of the states.",
    x="Date",
    y="degrees Celcius") + 
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
```

Checking for multicolinearity between min, max temperature, and wind speed:
```{r}
ggpairs(Covid19Climate, columns=c("Max", "Min", "Wind"))
```

## Statistical Analysis and Interpretation:

Test for correlation between daily wind speed, minimum, maximum temperature and the number of positive test case.
```{r}
cor.test(Covid19Climate$Min, Covid19Climate$positiveCaseViral, method = "pearson")
cor.test(Covid19Climate$Max, Covid19Climate$positiveCaseViral, method = "pearson")
cor.test(Covid19Climate$Wind, Covid19Climate$positiveCaseViral, method = "pearson")
```

Conducting correlation test for all three variables we have the following values: t is the t-test statistic value, with maximum, minimum temperature and wind speed valued at t = 20, 16, -7 respectively. df is the degrees of freedom, uniformly valued at df= 3067, p-value is the significance level of the t-test, which is infinitesimally small.Confidence interval of the correlation coefficient at 95% value at respectively [0.315 0.377] for maximum temperature, [0.251, 0.316] for minimum temperature, [-0.1584 -0.0887] for wind speed. Sample estimates are the estimated correlation coefficient, which valued at 0.346 for maximum temperature, 0.284 for minimum temperature, -0.124 for wind speed.

```{r warning=FALSE, message=FALSE}
train_test <- split(Covid19Climate, sample(0:107, size=2, replace=F))
test <- as.data.frame(train_test[2])
train <- as.data.frame(train_test[1])
colnames(test) <- colnames(Covid19Climate)
colnames(train) <- colnames(Covid19Climate)

maxTempModel <- lm(positiveCaseViral ~ Max, data=Covid19Climate)
minTempModel <- lm(positiveCaseViral ~ Min, data=Covid19Climate)
WindModel <- lm(positiveCaseViral ~ Wind, data=Covid19Climate)

summary(maxTempModel)
summary(minTempModel)
summary(WindModel)
```

As conducted in the data exploration part, the three maximum temperature, minimum temperature and wind speed has significant multicolinearity with each other, therefore causing major problems if multivariate regression is conducted. Thus I have conducted 3 separate linear regression to test for separate correlation between each independent variable and the dependent variable. These 3 linear regression model respectively describes the relationship between the number of positive test case based on maximum temperature, minimum temperature, and wind speed of each states over the months. Running the summary function on all 3 models gives us the general view of all 3 models. The values are the residuals and its statistical values (mean, mode, median, min, max), estimated coefficients and its standard errors, the R-squared metric, the F-value. 

The stars (***) are the significance level, computed and displayed based on the p-value of the model. Based on the summary, we can see that all three models level 2 or 3 significance, indicating that it's unlikely that no relationship exist between the number of positive test cases and climate values (minimum temperature, maximum temperature, wind speed).

As for the estimated coefficients, this value represents the slope of the function as calculated by the regression. In our 3 models, the estimated coefficients for maximum temperature, minimum temperature, wind speed, are respectively 109117, 123043, and -851230. In other words, for every increase of 1 degrees Celcius of maximum temperature, the predicted number of infected increase by 109117. For every increase of 1 degrees Celcius of minimum temperature, the predicted number of infected increase by 123043. For every increase of 1 meters per second of wind speed, the predicted number of infected decrease by 851230.

As the name of the value suggest, estimated coefficients has its errors in the estimation. The standard Error is the measure of the variance in the coefficient. In the three models, the standard errors are respectively 28422, 28093, 306391. The value for maximum and minimum temperature is roughly 5 times smaller than the estimated coefficient itself. The value for wind speed, however, is up to half the estimated coefficients.

R-squared is a metric used to evaluate the fit precision of the model. The maximum temperature, minimum temperature and the wind speed model all respectively has R-squared value of 12.3%, 15.4%, 6.85%. These values means that the model explains very few of the variance in the observation around the line of best fit.

The F-value in these models are the results of tests where the null hypothesis is "all of the regression coefficients are equal to zero". Since the p-value of all 3 models are much less than the alpha level, the null hypothesis can be rejected. All the models have their F-statistic relatively larger than 1, given the size of the data, insinuating that there may be correlations between the independent variables and the number of positive test cases in the US.


```{r warning=FALSE, message=FALSE}
test <- test %>%
  mutate(predictionWindDiff =  abs(positiveCaseViral - predict(WindModel, test))) %>%
  mutate(predictionMinDiff =  abs(positiveCaseViral - predict(minTempModel, test))) %>%
  mutate(predictionMaxDiff =  abs(positiveCaseViral - predict(maxTempModel, test)))
```

Histograms representing difference between predicted number of positive cases and actual positive cases:
```{r warning=FALSE, message=FALSE}
ggplot(test, aes(test$predictionMaxDiff)) + 
  geom_histogram(fill="steelblue") +
  labs(
    title="Distribution of maximum temperature prediction difference",
    caption="Histogram represents the distribution of the maximum temperature model's prediction difference",
    x="Confirmed cases difference") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))

ggplot(test, aes(test$predictionMinDiff)) + 
  geom_histogram(fill="steelblue") +
  labs(
    title="Distribution of minimum temperature prediction difference",
    caption="Histogram represents the distribution of the minimum temperature model's prediction difference",
    x="Confirmed cases difference") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))

ggplot(test, aes(test$predictionWindDiff)) + 
  geom_histogram(fill="steelblue") +
  labs(
    title="Distribution of wind speed prediction difference",
    caption="Histogram represents the distribution of the wind speed model's prediction difference",
    x="Confirmed cases difference") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
```



## Conclusions:

The data suggest that there exist a correlation between each independent variable of maximum temperature, minimum temperature, and wind speed to the dependent variable. However, there exist limitations in this data and the regression model. Because the awareness of Covid-19 not starts at around March, as shown the in sudden increase of positive test case graph, the data gathering process does not fully start until the beginning of April. This leads to incomplete data in between the months of January and March. The climate data set is gathered by many stations in a single state. The processing step takes the mean value of all the stations observation to represent the state's climate observations, which may or may not truely represent the state's climate situation. This may effect the regression analysis's accuracy because the data itself may or may not truely represent the situation. Another limitation in the regression and choice of exploratory variable is that these exploratory variables have high correlation with each other. In the future, the research will be looking for different exploratory or independent variable that represents the climate and how it effects the spread of Covid-19.

Reference that is relevant and supports my insights[8]: 

Xie, Jingui, and Yongjian Zhu. “Association between Ambient Temperature and COVID-19 Infection in 122 Cities from China.” Science of The Total Environment, Elsevier, 30 Mar. 2020, www.sciencedirect.com/science/article/pii/S0048969720317149

## Works Cited:

>1. “How Coronavirus Spreads.” Centers for Disease Control and Prevention, Centers for Disease Control and Prevention, www.cdc.gov/coronavirus/2019-ncov/prevent-getting-sick/how-covid-spreads.html
This source is published by the CDC as a way to inform the public with basic information about COVID 19. Specifically in this source, it gives facts about how COVID-19 spreads as well as stating some researches claiming various theories about factors that affect the spread of COVID-19. The claims made in this source serves as the foundation of this research.

>2. Pirouz, Behzad, et al. “Relationship between Average Daily Temperature and Average Cumulative Daily Rate of Confirmed Cases of COVID-19.” MedRxiv, Cold Spring Harbor Laboratory Press, 1 Jan. 2020, www.medrxiv.org/content/10.1101/2020.04.10.20059337v3

>Prata, David N., et al. “Temperature Significantly Changes COVID-19 Transmission in (Sub)Tropical Cities of Brazil.” Science of The Total Environment, Elsevier, 25 Apr. 2020, www.sciencedirect.com/science/article/pii/S004896972032379
Xie, Jingui, and Yongjian Zhu. “Association between Ambient Temperature and COVID-19 Infection in 122 Cities from China.” Science of The Total Environment, Elsevier, 30 Mar. 2020, www.sciencedirect.com/science/article/pii/S0048969720317149
These sources are studies with a closely resembling research question with my research, however, it used a different dependent variable that corresponds to the actual spread of COVID-19. The dependent variable these groups used was confirmed cases and is conducted in China and Brazil and all the countries’ cities. These sources is used as a comparison on how the two different dependent variables might correspond to the actual spread of COVID-19.

>3. Roser, Max, et al. “Coronavirus Pandemic (COVID-19) - Statistics and Research.” Our World in Data, 4 Mar. 2020, www.ourworldindata.org/coronavirus
This source is actually a data set but it explains how each variable affects the spread of COVID-19. Those explanations include why it is more effective and precise to measure the spread of COVID-19 in terms of tests per confirmed case.

>4. “Our Data.” The COVID Tracking Project, www.covidtracking.com/data
This source is the data set used to obtain all the COVID-19 data, including the dependent variable tests per confirmed case.

>5. Eeemonts. “Weather/Climate Data for JHU's COVID19 Sites.” Kaggle, 24 Sept. 2020, www.kaggle.com/eeemonts/weatherclimate-data-covid19
This source is the data set used to obtain the independent climate variables.

>6. “Data Definitions.” The COVID Tracking Project, www.covidtracking.com/about-data/data-definitions
This source is the metadata for the climate variable data set.

>7. Grossman, Adam. Dark Sky Blog, 1 Aug. 2020, www.blog.darksky.net 
This source is the document about Dark Sky, the company responsible for collecting the weather data for the climate data set.

>8. Xie, Jingui, and Yongjian Zhu. “Association between Ambient Temperature and COVID-19 Infection in 122 Cities from China.” Science of The Total Environment, Elsevier, 30 Mar. 2020, www.sciencedirect.com/science/article/pii/S0048969720317149

